---
title: "*Colias* Redundancy Analysis"
author: "Lily Durkee"
date: "2024-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Upload raw VCF file, sample 10,000 SNPs

```{r}
# Install and load the adegenet package
#install.packages("adegenet")
library(adegenet)
library(vegan)

# Set the path to your PLINK .raw file (without the file extension)
raw_vcf <- "Data/colias.4x.merged_gatk.rm.relate.SNP.filtered_gatkVQSR2.PASS.8miss.recode.vcf.gl_impute4.1.raw"

# Read the .raw file into R
colias_vcf <- read.PLINK(raw_vcf)

# view the number of NAs
sum(is.na(colias_vcf))

# isolate sample names
samples_all <- data.frame(ID=pop(colias_vcf))

# number of SNPs
snps_n <- nLoc(colias_vcf)

# sample 100k
set.seed(123)  # for reproducibility
snps_select <- sample(snps_n, 100000)
colias_vcf.sub <- colias_vcf[,snps_select]
```
## Upload (lat,long) coordinates for butterflies
```{r}
#install.packages("elevatr")
library(elevatr)
library(raster)
library(tidyverse)
library(geodata)
library(sf)

# this file is updated to include correct sample IDs
all.colias <- read.csv("Data/Colias-samples-2022-all.csv")


#bioclim_vars <- getData("worldclim", var="bio", res=.5, lat = all.colias$y, lon = all.colias$x)

#bioclim8 <- raster("wc0.5/bio8_12.bil")
#bioclim13 <- raster("wc0.5/bio13_12.bil")

#### merge with sample IDs: colias_samples ####
samples_all <- data.frame(ID=pop(colias_vcf))
colias_samples1 <- merge(samples_all, all.colias, by="ID", all=F)

latlong <- colias_samples1 %>% dplyr:: select(x, y)
```


## Load bioclim variables

```{r}
library(tidyverse)
colias_clim1 <- read.csv("colias_bioclim.csv") %>% dplyr::select(-X)
latlong <- colias_clim1 %>% dplyr:: select(x, y)
```

## Load elevation using package `elevatr`

```{r}
library(elevatr)
#library(raster)
library(tidyverse)
#library(geodata)
library(sf)

# #### extract elevation data using elevatr ####
crs <- st_crs("+proj=longlat +datum=WGS84")

# Convert your elevation.df data frame to an sf object
elevation.sf <- st_as_sf(latlong, coords = c("x", "y"), crs = crs)

elevation <- get_elev_point(elevation.sf)
#### final file ####

#colias_clim1 <- cbind(colias_samples1, data.frame(tmax=tmax, precip=prec, wind=wind, srad=srad))
colias_clim.final <- cbind(colias_clim1, elevation)

#write.csv(colias_clim, "colias_bioclim.csv")
```


## Run the RDA

```{r}
library(vegan) # to run the RDA
library(psych) # for visualizing correlations
#library(viridis) # for colors

# ## look at samples in VCF
# # use the pop() function to retrieve the sample names
# sample_names <- pop(colias_vcf)
# 
# ## compare samples in each file
# list1 <- print(sample_names)
# #colias_samples$ID <- as.factor(colias_samples$ID)
# list2 <- colias_samples[,1]
# 
# setdiff(list1, list2)


## will need environment data for every point, in the same order
identical(as.character(samples_all[,1]), colias_clim.final[,1]) 

#TRUE!

#colias_clim$ID <- as.character(colias_clim$ID)

# create ENV file that has the environmental variables to include
colias_env.all <- colias_clim.final %>% dplyr::select(c(elevation, tmax, precip, srad, wind))
pairs.panels(colias_env.all, scale=T)

colias_env <- colias_clim.final %>% dplyr::select(c(elevation, precip, srad))
pairs.panels(colias_env, scale=T)
#png("colias_bioclim.corr_final.png")

#### run the RDA on all the data - file too big, takes forever ####
#colias_rda <- rda(colias_vcf ~ ., data=colias_env, scale=T)
#colias_rda

#### run the RDA on a subset of 100k SNPs ####
colias_rda.sub <- rda(colias_vcf.sub ~ ., data=colias_env, scale=T)
colias_rda.sub

# R squared
RsquareAdj(colias_rda.sub)
screeplot(colias_rda.sub)

# plot the RDA model
plot(colias_rda.sub)
# visualize RDA axes

# check for significance / this takes forever
#signif.full <- anova.cca(colias_rda.sub, parallel=getOption("mc.cores")) # default is permutation=999
signif.full

# variance inflation factor - tells if variables are correlated
vif.cca(colias_rda.sub)

#### RDA plots ####
# try a plot
plot(colias_rda.sub, scaling=3)

# more informative plots
colias_env$site <- colias_clim1$site
unique(colias_clim1$site)
levels(colias_env$site) <- as.fac("Babbit Gulch","Carpenter Ranch","Dumont Lake","Fountain Valley","Harbison Meadows","High Trails Ranch","Kebler Pass","D Loukonen Farm","Little Laramie","Schreiner Farm","North Fork Trail","Orchard Creek Ranch","Spring Canyon Park","Soapstone Prairie")
eco <- colias_env$site
bg <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd",
                     "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
                     "#aec7e8", "#ffbb78", "#98df4a", "#ff9896")


# match site names to colors
site_colors <- bg[match(eco, levels(eco))]
png("Colias_RDA_attempt1-axis1&2.png", width = 10, height = 8, units = "in", res = 350)

plot(colias_rda.sub, type="n", scaling=3, cex.axis=1.2, cex.lab=1.2)
points(colias_rda.sub, display="species", pch=4, cex=.7, col="gray32", scaling=3)  # the SNPs
points(colias_rda.sub, display="sites", pch=21, cex=2, scaling=3, bg=site_colors)  # the butterflies
text(colias_rda.sub, scaling=3, display="bp", col="#0868ac", cex=1.5)  # the predictors
legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)

dev.off()

## axis 1 & 3
png("Colias_RDA_attempt1-axis1&3.png", width = 12, height = 12, units = "in", res = 350)

plot(colias_rda.sub, type="n", scaling=2, cex.axis=1.2, cex.lab=1.2, choices=c(1,3))
points(colias_rda.sub, display="species", pch=4, cex=.7, col="gray32", scaling=3)  # the SNPs
points(colias_rda.sub, display="sites", pch=21, cex=1.5, scaling=2, bg=site_colors)  # the butterflies
text(colias_rda.sub, scaling=3, display="bp", col="#0868ac", cex=1)  # the predictors
legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)

dev.off()
```

## Identifying candidate SNPs
```{r}
#### run the RDA on ALL 1.4mil SNPs ####
colias_rda <- rda(colias_vcf ~ ., data=colias_env, scale=T)
colias_rda

load.rda <- scores(colias_rda, choices=c(1:3), display="species")

# relatively normal distributions 
layout(matrix(1:3, nrow = 1))
hist(load.rda[,1], main="Loadings on RDA1")
hist(load.rda[,2], main="Loadings on RDA2")
hist(load.rda[,3], main="Loadings on RDA3")

# we are interested in SNPs at the tails of the distributions
# this function will be able to detect outlier SNPs
outliers <- function(x,z){
  lims <- mean(x) + c(-1, 1) * z * sd(x)     # find loadings +/-z sd from mean loading     
  x[x < lims[1] | x > lims[2]]               # locus names in these tails
}

# candidate loci
cand1 <- outliers(load.rda[,1],3) # 651
cand2 <- outliers(load.rda[,2],3) # 214
cand3 <- outliers(load.rda[,3],3) # 214

# number of candidate loci identified
ncand <- length(cand1) + length(cand2) + length(cand3)
ncand 
 
#1,079 with 100k SNPs
# 23,618 with all SNPs

# now, we will organize our results in one data frame with axis, SNP name, & correlation with each predictor
cand1 <- cbind.data.frame(rep(1,times=length(cand1)), names(cand1), unname(cand1))
cand2 <- cbind.data.frame(rep(2,times=length(cand2)), names(cand2), unname(cand2))
cand3 <- cbind.data.frame(rep(3,times=length(cand3)), names(cand3), unname(cand3))

colnames(cand1) <- colnames(cand2) <- colnames(cand3) <- c("axis","snp","loading")

cand <- rbind(cand1, cand2, cand3)
cand$snp <- as.character(cand$snp)

head(cand)

# we will now put them in a single data frame with the 3 env predictors
foo <- matrix(NA, nrow=(ncand), ncol=3)  # 3 columns for 3 predictors
colnames(foo) <- c("elevation", "precip", "srad")

for (i in 1:length(cand$snp)) {
  nam <- cand[i,2]
  snp.gen <- as.matrix(colias_vcf)[,nam]
  foo[i,] <- apply(dplyr::select(colias_env, c(elevation, precip, srad)), 2, 
                   function(x) cor(x,snp.gen))
}

cand <- cbind.data.frame(cand,foo)  
head(cand)

# 1075 total duplicates
length(cand$snp[duplicated(cand$snp)])

# axis 1
foo <- cbind(cand$axis, duplicated(cand$snp)) 
table(foo[foo[,1]==1,2]) 

# axis 2
table(foo[foo[,1]==2,2])

# axis 3
table(foo[foo[,1]==3,2]) 

cand_no.dup <- cand[!duplicated(cand$snp),] # remove duplicate detections
# only 4 left

# see which environmental vars they are most correlated with

for (i in 1:length(cand_no.dup$snp)) {
  bar <- cand_no.dup[i,]
  cand_no.dup[i,7] <- names(which.max(abs(bar[3:6]))) # gives the variable
  cand_no.dup[i,8] <- max(abs(bar[3:6]))              # gives the correlation
}

colnames(cand_no.dup)[7] <- "predictor"
colnames(cand_no.dup)[8] <- "correlation"

table(cand_no.dup$predictor) 

# for 100k SNPs
# elevation   loading      srad 
#         1         1         2 
```

# APPENDIX - Climate variable upload

## Upload climate data - BioClim - these data are now in file `colias_bioclim.csv`
```{r}
library(raster)

# List of months
months <- c("06", "07", "08")

#### WIND ####

# Initialize an empty raster stack to store the monthly wind rasters
wind_rasters <- stack()

# Loop through each month
for (month in months) {
  # Construct the file name for the wind raster
  file_name <- paste0("Data/bioclim-wind/wc2.1_30s_wind_", month, ".tif")
  
  # Load the wind raster for the current month
  wind_raster <- raster(file_name)
  
  # Add the wind raster to the raster stack
  wind_rasters <- addLayer(wind_rasters, wind_raster)
}

# Calculate the mean across the raster stack to get the average wind for June, July, and August
mean_wind <- mean(wind_rasters)

## extract points
wind <- extract(mean_wind, latlong)

#### TMAX ####

# Initialize an empty raster stack to store the monthly wind rasters
tmax_rasters <- stack()

# Loop through each month
for (month in months) {
  # Construct the file name for the wind raster
  file_name <- paste0("Data/bioclim-tmax/wc2.1_30s_tmax_", month, ".tif")
  
  # Load the tmax raster for the current month
  tmax_raster <- raster(file_name)
  
  # Add the tmax raster to the raster stack
  tmax_rasters <- addLayer(tmax_rasters, tmax_raster)
}

# Calculate the mean across the raster stack to get the average wind for June, July, and August
mean_tmax <- mean(tmax_rasters)

## extract points
tmax <- extract(mean_tmax, latlong)

#### PRECIPITATION ####

# Initialize an empty raster stack to store the monthly precip rasters
prec_rasters <- stack()

# Loop through each month
for (month in months) {
  # Construct the file name for the precip raster
  file_name <- paste0("Data/bioclim-prec/wc2.1_30s_prec_", month, ".tif")
  
  # Load the precip raster for the current month
  prec_raster <- raster(file_name)
  
  # Add the precip raster to the raster stack
  prec_rasters <- addLayer(prec_rasters, prec_raster)
}

# Calculate the mean across the raster stack to get the average wind for June, July, and August
mean_prec <- mean(prec_rasters)

## extract points
prec <- extract(mean_prec, latlong)


#### SOLAR RADIATION ####

# Initialize an empty raster stack to store the monthly srad rasters
srad_rasters <- stack()

# Loop through each month
for (month in months) {
  # Construct the file name for the srad raster
  file_name <- paste0("Data/bioclim-srad/wc2.1_30s_srad_", month, ".tif")
  
  # Load the srad raster for the current month
  srad_raster <- raster(file_name)
  
  # Add the srad raster to the raster stack
  srad_rasters <- addLayer(srad_rasters, srad_raster)
}

# Calculate the mean across the raster stack to get the average wind for June, July, and August
mean_srad <- mean(srad_rasters)

## extract points
srad <- extract(mean_srad, latlong)
```

## Upload climate data - Adapt West  - *not granular enough*

```{r}
# #library(raster)
# 
# # List of variable names
# variables <- c("PPT", "Tmax")
# 
# # List of months
# months <- c("06", "07", "08")


# ## extract PPT values for June-August
# PPT06.ras <- raster("Normal_1991_2020_monthly/Normal_1991_2020_PPT06.tif")
# PPT07.ras <- raster("Normal_1991_2020_monthly/Normal_1991_2020_PPT07.tif")
# PPT08.ras <- raster("Normal_1991_2020_monthly/Normal_1991_2020_PPT08.tif")
# 
# PPT06 <- extract(PPT06.ras, latlong)
# PPT07 <- extract(PPT07.ras, latlong)
# PPT08 <- extract(PPT08.ras, latlong)
# 
# ## extract Tmax values for June-August
# Tmax06 <- raster("Normal_1991_2020_monthly/Normal_1991_2020_Tmax06.tif")
# Tmax07 <- raster("Normal_1991_2020_monthly/Normal_1991_2020_Tmax07.tif")
# Tmax08 <- raster("Normal_1991_2020_monthly/Normal_1991_2020_Tmax08.tif")
# 
# Tmax06 <- extract(Tmax06, latlong)
# Tmax07 <- extract(Tmax07, latlong)
# Tmax08 <- extract(Tmax08, latlong)
# 
# colias_clim1 <- cbind(colias_samples1, data.frame(Tmax06=Tmax06, 
#                                                   Tmax07=Tmax07,
#                                                   Tmax08=Tmax08,
#                                                   PPT06=PPT06,
#                                                   PPT07=PPT07,
#                                                   PPT08=PPT08))
# 
# ## all of these values are the same :(

## try precipitation as snow

#pas.ras <- raster("Normal_1991_2020_bioclim/Normal_1991_2020_PAS.tif")
#pas <- extract(pas.ras, latlong)
## these are also the same value :(

# # Function to calculate the mean across months
# calculate_mean <- function(variable) {
#   # Initialize an empty list to store raster objects
#   raster_list <- list()
#   
#   # Load raster files for each month
#   for (month in months) {
#     filename <- paste0("Normal_1991_2020_monthly/Normal_1991_2020_", variable, month, ".tif")
#     raster_list[[month]] <- raster(filename)
#   }
#   
#   # Extract values for each raster at the given lat/long points
#   values_list <- lapply(raster_list, function(r) extract(r, latlong))
#   
#   # Calculate the mean for each set of values
#   mean_values <- lapply(values_list, function(x) {
#     print(dim(x))
#     apply(x, 2, mean, na.rm = TRUE)
#     })
#   
#   return(mean_values)
# }
# 
# # Iterate over each variable
# for (variable in variables) {
#   # use the function above on each variable
#   mean_values <- calculate_mean(variable)
#   
#   # create a raster with the mean values
#   #mean_raster <- raster(matrix(mean_values, ncol = ncol(values_list[[1]])))
# 
#   # save the mean raster to a new file
#   #writeRaster(mean_raster, filename = paste0("mean_", variable, "_June_August.tif"), format = "GTiff")
# }

```

## Extract climate data using dismo - *lots of errors*

```{r}
# library(dismo)
# # extract bioclim variables
# # bioclim_var8 <- data.frame(raster::extract(bioclim8, latlong))
# # bioclim_var13 <- data.frame(raster::extract(bioclim13, latlong))
# 
# colias_samples2 <- cbind(colias_samples1, bio8=bioclim_var8, bio18=bioclim_var13)
# 
# #all.colias$bio5_12 <- all.colias$bio5_12/10
# 
# colias_samples.final$bio8 <- colias_samples.final$raster..extract.bioclim8..latlong.
# colias_samples.final$bio13 <- colias_samples.final$raster..extract.bioclim13..latlong.

```
